The role of cognateness in native spoken word recognition

### 2024-07-25 manuscript/manuscript.tex
New version of introduction

\usepackage{booktabs}
\usepackage{caption}
\usepackage{longtable}
\usepackage{colortbl}
\usepackage{array}
\usepackage{tipa}
\usepackage{booktabs}
\DeclareTextFontCommand\textphonetic{\ipafont}
\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{english}
\ifPDFTeX
\else
\babelfont{rm}[]{Latin Modern Roman}
\fi
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={The role of cognateness in recognition of spoken words in a foreign language},
  pdfauthor={Gonzalo Garcia-Castro; Serene Siow; Kim Plunkett; Nuria Sebastian-Galles},
  pdflang={en},
  pdfkeywords={cognate, spoken word
recognition, phonology, bilingualism, non-native speech
processing, lexical access},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\usepackage{lineno}
\linenumbers
\newcommand{\runninghead}{A Preprint }
\renewcommand{\runninghead}{Cognateness and recognition of foreign
words }
\title{The role of cognateness in recognition of spoken words in a
foreign language}
\author{\textbf{Gonzalo
Garcia-Castro}~\orcidlink{0000-0002-8553-4209}\\Center for Brain and
Cognition\\Universitat Pompeu Fabra\\Barcelona,
08005\\\texttt{\href{mailto:gonzalo.garciadecastro@upf.edu}{gonzalo.garciadecastro@upf.edu}}\And\textbf{Serene
Siow}~\orcidlink{0000-0001-6482-2191}\\Department of Experimental
Psychology\\University of Oxford\\Oxford, OX2
6GG\\\texttt{\href{mailto:siow.serene@gmail.com}{siow.serene@gmail.com}}\And\textbf{Kim
Plunkett}~\orcidlink{0000-0003-0216-7480}\\Department of Experimental
Psychology\\University of Oxford\\Oxford, OX2
6GG\\\texttt{\href{mailto:kim.plunkett@psy.ox.ac.uk}{kim.plunkett@psy.ox.ac.uk}}\And\textbf{Nuria
Sebastian-Galles}~\orcidlink{0000-0001-6938-2498}\\Center for Brain and
Cognition\\Universitat Pompeu Fabra\\Barcelona,
08005\\\texttt{\href{mailto:nuria.sebastian@upf.edu}{nuria.sebastian@upf.edu}}}
\date{}
\begin{document}
\maketitle
\begin{abstract}
Understanding spoken words in a non-native language is easier when their
translations into the listener's native language are phonologically
similar. Phonological similarity plays a more important role in the
language performance of low-proficiency bilinguals, compared to that of
high-proficiency bilinguals. This suggests that low-proficiency
bilinguals may points to the latter as relying more strongly on
cross-language word-to-word links (lexical route), while
high-proficiency bilinguals rely more on word-concept associations
(conceptual route) than on cross-language word-to-word links (lexical
route). Disentangling bilinguals' reliance on the conceptual and lexical
and conceptual routes during translation is challenging, given that both
sources of information are present during spoken word comprehension to
some extent in both high- and low-proficiency bilinguals. In this study,
we tested English and Spanish native speakers in a translation
elicitation task, in which they listened to words in an unfamiliar
language (Spanish or Catalan for English natives, and Catalan for
Spanish natives), and then had to translate them to their native
language. Given their lack of previous knowledge on the unfamiliar
tested language, phonological similarity between the presented words and
their correct translations was the only cue available for participants
to succeed in the task. This allowed us to explore the informativeness
of the lexical route for word translation, in the absence of information
from the conceptual route. Our results indicate that participants
benefited from the phonological similarity between translation
equivalents: the more similar the word they listened to and its
translational equivalent, the higher the probability of a correct
translation. These results point to cross-language word-to-word links as
providing rich information during word translation.
\end{abstract}
{\bfseries \emph Keywords}
\def\sep{\textbullet\ }
cognate \sep spoken word
recognition \sep phonology \sep bilingualism \sep non-native speech
processing \sep 
lexical access

@@

Learning a new language is a non-trivial task, a large part of which is
acquiring a new set of word-forms to correspond to existing semantic
representations in their lexicon. Imagine that a Spanish listener hears
English for the first time and encounters the word /\textipa{"dO:}/
(\emph{door}). Their knowledge of the phonological form in their native
language is of little help: /\textipa{"pweR.ta}/ and /\textipa{"dO:}/ do
not share any phonemes. It would come as a surprise if the listener was
able to translate the English word successfully without having first
learnt the word. However, there are some elements in language that may
help the learner in acquiring words in a new language. Many languages
share some degree of similarity. This may be due to typological
closeness and/or socio-historical events involving the speakers of these
languages (e.g., migration, social contact). For example, Romance
languages such as Spanish and Catalan share many form-similar
translation equivalents (also known as cognates)
(\citeproc{ref-schepens2012distributions}{Schepens et al., 2012}), as in
the case of \emph{puerta} and \emph{porta} (\emph{door} in Spanish and
Catalan, respectively). Given no prior knowledge of either language, a
Spanish native speaker is likely to be much more successful at correctly
translating Catalan \emph{porta} than English \emph{door}, due to the
phonological and orthographical similarity of the former to the
equivalent in their native language. Second language learners have been
found to learn cognates more easily than non-cognates
(\citeproc{ref-hall2002automatic}{Hall, 2002};
\citeproc{ref-otwinowska2019more}{Otwinowska \& Szewczyk, 2019}). This
type of cross-linguistic similarity has also been extremely informative
for the study of bilingual lexical processing: words that share high
form-similarity with their translations are processed faster and more
accurately than words sharing lower phonological similarity. This
phenomenon has provided strong evidence in favour of the language-non
selective hypothesis of bilingual lexical access, which states that
bilinguals activate lexical representations in both languages, even
during monolingual situations. This facilitation effect has been
reported for comprehension (\citeproc{ref-dijkstra2010cross}{Dijkstra et
al., 2010}; \citeproc{ref-midgley2011effects}{Midgley et al., 2011};
\citeproc{ref-thierry2007brain}{Thierry \& Wu, 2007}), production
(\citeproc{ref-costa2000cognate}{Costa et al., 2000}), learning
(\citeproc{ref-de2000hard}{De Groot \& Keijzer, 2000};
\citeproc{ref-elias2022crosslanguage}{Elias \& Degani, 2022};
\citeproc{ref-lotto1998effects}{Lotto \& De Groot, 1998};
\citeproc{ref-valente2018does}{Valente et al., 2018}), and translation
(\citeproc{ref-christoffels2006memory}{Christoffels et al., 2006}).

Event counting on this facilitation effect of form-similarity, listening
to non-native speech is costlier than listening to native speech, even
for highly proficient bilinguals, and especially in acoustically adverse
situations (\citeproc{ref-lecumberri2010non}{Lecumberri et al., 2010};
\citeproc{ref-takata1990english}{Takata \& Nábělek, 1990}). One of
potential source of this increased difficulty is the partial mismatch
between the phonology of the native and the non-native languages: some
acoustic features embedded in the non-native speech signal do not
overlap perfectly with any phonemic category in the listener's native
language. For example, imagine a Spanish native with no previous
familiarity with any other language listening to French for the first
time. This listener may encounter the word /\textipa{"pOKt}/
(\emph{porte}), translation of door in French. When comparing the
phonemic transcription to the transcription of the Spanish word
/\textipa{"pweR.ta}/ (\emph{puerta}), one may be led to infer that these
two words do not have much overlap. But one thing to note is that the
voiced uvular fricative consonant /\textipa{K}/ and the open-mid back
rounded vowel /\textipa{O}/, which do not exist in Spanish as phonemic
categories, can be perceived as allophonic variations the native
(Spanish) phonemes /\textipa{R}/ and /o/. A perceptual assimilation
model has been proposed to account for how speakers may integrate
non-native phonemes into their native phonemic categories based on
goodness-of-fit (\citeproc{ref-best1988examination}{Best et al., 1988},
\citeproc{ref-best2001discrimination}{2001}). Although phonetic mismatch
between the non-native word and its translation can have a noticeable
toll on comprehension Cutler et al.
(\citeproc{ref-cutler2004patterns}{2004}), the Spanish monolingual may
still be able to activate the phonological representation of
/\textipa{"pweR.ta}/ (\emph{puerta}) from hearing the word-form
/\textipa{"pOKt}/ (\emph{porte}), ultimately allowing them to access the
right semantic representation.

While phonological similarity across translation equivalents can help
with speech recognition, phonological similarity between words with
different meanings can hinder the recognition process. In word
retrieval, the presence of phonological neighbours (word-forms that
differ from each other in one phoneme) may interfere with the activation
and selection of the appropriate lexical representation in the native
language. Words with denser phonological neighbourhoods are recognised
more slowly and less accurately than words with sparser phonological
neighbourhoods (\citeproc{ref-dufour2003lexical}{Dufour \& Peereman,
2003}; \citeproc{ref-goldinger1989priming}{Goldinger et al., 1989};
\citeproc{ref-hamburger1996phonological}{Hamburger \& Slowiaczek, 1996};
\citeproc{ref-luce1990similarity}{Luce et al., 1990};
\citeproc{ref-luce1998recognizing}{Luce \& Pisoni, 1998}). This is
especially true if such neighbours share higher phonological similarity
with the presented word, or are lexically more frequent than the target
translation (\citeproc{ref-luce1998recognizing}{Luce \& Pisoni, 1998}).
The inhibitory effect of phonological neighbourhoods has also been
reported across languages. For instance, Weber \& Cutler
(\citeproc{ref-weber2004lexical}{2004}) observed that bilinguals
displayed lexical competition effects from words in their native
language when performing a word-referent matching task in their L2. The
assimilation of non-native phonemes into native categories
(\citeproc{ref-best1988examination}{Best et al., 1988},
\citeproc{ref-best2001discrimination}{2001}) was used to explain how
lexical competitors in the native language can be activated even when
the target words contain non-native phonemes. In the case of non-native
speech recognition, the presence of cross-linguistic pairs which are
phonologically similar but differ in meaning (e.g., false friends) may
act as distractors during lexical access, obstructing the selection of
the appropriate target translation in the listener's lexicon. For
instance, Otwinowska \& Szewczyk
(\citeproc{ref-otwinowska2019more}{2019}) reported that false friends
were disadvantaged relative to non-cognates by Polish second language
learners, in contrast to cognates which were known better. It is
therefore important to investigate the joint effect of both cognates and
false friends when investigating the effect cross-linguistic
phonological similarity has on word recognition in a foreign language.
This could shed light on available strategies and challenges associated
with the early stages of second language acquisition.

In the present study, we investigate the extent to which monolinguals
can exploit phonological similarity between translation equivalents when
listening to words in an unfamiliar language, and how it interacts with
competition from phonological neighbours in the native language. We used
a translation elicitation task in which participants listened to words
from an unfamiliar language, and then tried to guess their translation
in their native language. We will henceforth refer to the
auditory-presented words heard by participants on each trial as
presented words, and the correct translation for the presented words as
target translations. We explored listeners' reliance on phonological
similarity to succeed in the task by manipulating the amount of
phonological similarity between the presented words and their target
translations. Since participants were unfamiliar with the presented
language, they could only use phonological similarity between the
presented language and their native language to successfully translate
the words. We therefore predicted that participants' performance should
increase when the translation pairs are phonologically more similar. We
also predicted that there is a minimum threshold of phonological
similarity to be sufficient for translation.

We further investigated the effect of phonological competitors. Even if
the presented word and its target translation share high phonological
similarity, participants may still provide incorrect translations if the
presented word also shares high phonological similarity with other words
in the native language. The effect is predicted to be especially strong
if any of the competitors is more lexically frequent than the target
translation. We hypothesised that the size of the facilitatory role of
phonological similarity would be inversely proportional to the amount of
higher-frequency phonological neighbours of the presented word in the
target language. For instance, one could expect English participants to
incorrectly translate the Spanish word \emph{botón} as \emph{bottom}
instead of as its correct translation \emph{button}, due to combined
effect of the close phonological similarity between \emph{botón} and
\emph{bottom} along with the high lexical frequency of \emph{bottom}
relative to \emph{button}. To test this prediction, we developed a
lexical frequency-dependent measure of cross-language phonological
neighbourhood density, in which a neighbour is counted only if it is
higher frequency and is one phoneme apart from the presented word. If
the phonological neighbourhood density of the target translation affects
participants' performance negatively, this would suggest that
competitors in the native language affect recognition of non-native
words during recognition of foreign speech.We conducted a series of
three experiments to test these predictions.

@@

@@

@@

\bibitem[\citeproctext]{ref-hall2002automatic}
Hall, C. J. (2002). \emph{The automatic cognate form assumption:
Evidence for the parasitic model of vocabulary development}.

@@

@@

\bibitem[\citeproctext]{ref-lecumberri2010non}
Lecumberri, M. L. G., Cooke, M., \& Cutler, A. (2010). Non-native speech
perception in adverse conditions: {A} review. \emph{Speech
Communication}, \emph{52}(11-12), 864--886.

@@

@@

@@

\bibitem[\citeproctext]{ref-takata1990english}
Takata, Y., \& Nábělek, A. K. (1990). English consonant recognition in
noise and in reverberation by {Japanese} and {American} listeners.
\emph{The Journal of the Acoustical Society of America}, \emph{88}(2),
663--666.

>>>>

\usepackage{booktabs}
\usepackage{caption}
\usepackage{longtable}
\usepackage{colortbl}
\usepackage{array}
\usepackage{tipa}
\usepackage{booktabs}
\DeclareTextFontCommand\textphonetic{\ipafont}
\usepackage[utf8]{inputenc}
\usepackage{CJKutf8}
\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{english}
\ifPDFTeX
\else
\babelfont{rm}[]{Latin Modern Roman}
\fi
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={The role of cognateness in native spoken word recognition},
  pdfauthor={Gonzalo Garcia-Castro; Serene Siow; Kim Plunkett; Nuria Sebastian-Galles},
  pdflang={en},
  pdfkeywords={cognate, spoken word
recognition, phonology, bilingualism, non-native speech
processing, lexical access},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\usepackage{lineno}
\linenumbers
\newcommand{\runninghead}{A Preprint }
\renewcommand{\runninghead}{Cognateness and recognition of foreign
words }
\title{The role of cognateness in native spoken word recognition}
\author{\textbf{Gonzalo
Garcia-Castro}~\orcidlink{0000-0002-8553-4209}\\Center for Brain and
Cognition\\Universitat Pompeu Fabra\\Barcelona,
08005\\\texttt{\href{mailto:gonzalo.garciadecastro@upf.edu}{gonzalo.garciadecastro@upf.edu}}\And\textbf{Serene
Siow}~\orcidlink{0000-0001-6482-2191}\\Department of Experimental
Psychology\\University of Oxford\\Oxford, OX2
6GG\\\texttt{\href{mailto:siow.serene@gmail.com}{siow.serene@gmail.com}}\And\textbf{Kim
Plunkett}~\orcidlink{0000-0003-0216-7480}\\Department of Experimental
Psychology\\University of Oxford\\Oxford, OX2
6GG\\\texttt{\href{mailto:kim.plunkett@psy.ox.ac.uk}{kim.plunkett@psy.ox.ac.uk}}\And\textbf{Nuria
Sebastian-Galles}~\orcidlink{0000-0001-6938-2498}\\Center for Brain and
Cognition\\Universitat Pompeu Fabra\\Barcelona,
08005\\\texttt{\href{mailto:nuria.sebastian@upf.edu}{nuria.sebastian@upf.edu}}}
\date{}
\begin{document}
\maketitle
{\bfseries \emph Keywords}
\def\sep{\textbullet\ }
cognate \sep spoken word
recognition \sep phonology \sep bilingualism \sep non-native speech
processing \sep 
lexical access

@@

When some German speakers listened to the song \emph{The Power} by SNAP!
(\emph{Coyote Ugly}, 2000), many of them misheard the line ``I've got
the power'' as \emph{Agatha Bauer!} When Michael Jackson vocalised
``ma-ma-coo-sah'' in his song \emph{Wanna Be Startin' Somethin'}
(\emph{Thriller}, 1982), some Dutch speakers understood \emph{mama
appelsap} {[}mama applejuice{]}. When Spanish speakers listen to the
line ``Circumvent your thick ego'' from the song \emph{Pictures} by
System of a Down (\emph{Steal this Album!}, 2002), they tend to hear
\emph{Sácame de aquí, King-Kong} {[}take me out of here, King-Kong{]}.
When Japanese speakers listen to the line ``I want to hold your hand''
in the homonym song by The Beatles (\emph{Meet The Beatles}, 1964), they
often mishear \emph{Aho na hōnyōhan} {[}stupid urinator{]}. Outrageous
as these examples might sound to speakers of other languages, many
readers may know a few cases in their own native language. This auditory
illusion can feel quite real, and often inevitable after the first time
it happens. In Japanese, this phenomenon takes the name \emph{Soramimi}
(lit. ``empty ear''), after the \emph{Soramimi Hour}---a section of the
program \emph{Tamori Club} hosted by comedian Morita Kazuyoshi
(Tamori)---in which instances of this illusion were presented with
comedy purposes. Soramimi is a particular case of \emph{homophonic
translation}: words or phrases in one language are translated into
similar-sounding words and phrases in a different language, without
necessarily preserving the meaning. This phenomenon extends beyond song
lyrics, and may occur anytime when listening to non-native speech. For
instance, when listening to ``You know me?'' (\textipa{/ju: n@U mi:}/)
in English, Japanese-native listeners tend to understand \emph{yunomi}
(/\textipa{jUnomi}/), which is a type of Japanese teacup. Overall, the
occurence of Soramimi---and more generally, homophonic
translations---suggests that speech in an unfamiliar language has the
potential to activate lexical representations in the native language.

Although homophonic translations do not necessarily take place with the
preservation of meaning---which often results in outrageous
mistranslations like the ones aforementioned---homophonic translations
sometimes may lead to the activation of the intended meaning (i.e., a
correct translation). For example, imagine a Spanish native with no
previous familiarity with any other language listening to French for the
first time. This listener may encounter the word /\textipa{"pOKt}/
(\emph{porte}), translation of door in French. When comparing the
phonemic transcription of the French to that of its Spanish translation
/\textipa{"pweR.ta}/ (\emph{puerta}), one may be led to infer that these
two words do not have much overlap. But one thing to note is that the
voiced uvular fricative consonant /\textipa{K}/ and the open-mid back
rounded vowel /\textipa{O}/, which do not exist in Spanish as phonemic
categories, can be perceived as allophonic variations the native
(Spanish) phonemes /\textipa{R}/ and /o/, as described by assimilation
models of phoneme perception (\citeproc{ref-best1988examination}{Best et
al., 1988}, \citeproc{ref-best2001discrimination}{2001}). Although this
phonetic mismatch between the non-native word and its translation can
have a noticeable toll on comprehension Cutler et al.
(\citeproc{ref-cutler2004patterns}{2004}), the Spanish listener may
still be able to activate the phonological representation of
/\textipa{"pweR.ta}/ (\emph{puerta}) from hearing the word-form
/\textipa{"pOKt}/ (\emph{porte}), ultimately allowing them to access the
correct semantic representation. Given enough phonological similarity
between the non-native word and its translation in the native language,
a naïve listener may succeed at word comprehension when listening to an
unfamiliar language. Such phonological similarity between translation
equivalents---known as \emph{cognateness}---is common across many
languages, and is often due to typological closeness and/or
socio-historical events involving the speakers of these languages (e.g.,
migration, social contact). For example, Romance languages such as
Spanish and Catalan share many form-similar translation equivalents
(\citeproc{ref-schepens2012distributions}{Schepens et al., 2012}), as in
the case of \emph{puerta} and \emph{porta} (\emph{door} in Spanish and
Catalan, respectively). Given no prior knowledge of either language, a
Spanish native speaker is likely to be much more successful at correctly
translating Catalan \emph{porta} than English \emph{door}, due to the
phonological and orthographic similarity of the former to the equivalent
in their native language.

While Soramimi, and more generally homophonic translation, are popular
and frequent phenomena, the psycholinguistic literature on the topic is
relatively scarce, and has mostly focused on the phonological variables
involved in the occurence of Soramimi. For instance, Otake
(\citeproc{ref-otake2007interlingual}{2007}) analysed 194 instances of
Soramimi, broadcasted between 1992 and 2007 by the TV show Soramimi
hour. These instances consisted of homophonic translations of English
song lyrics to Japanese words and phrases. The vast majority of the
reported instances were phrases (96\%) and a few, to single words (4\%).
The phonetic features of the presented English lyrics were preserved
with relatively high variability, with some Japanese resulting words or
phrases sharing high overlap with the original English strings of
sounds, and some sharing very little overlap. When analysing the few
instances of homophonic translations of single words, the author
identified three phonological processes that might explain how Japanese
listeners reconstructed the English input to generate the Japanese
words: insertion (e.g., \emph{cry} /\textipa{kraI}/ to \emph{kurai},
/\textipa{kuRai}/ {[}dark{]}), deletion (e.g., \emph{go},
/\textipa{goU}/ to \emph{go}, /go/ {[}go{]}, and alternation (e.g.,
\emph{low}, /\textipa{loU}/ to \emph{rou}, /\textipa{Roo}/ {[}wax{]}).
This suggests that the reported homophonic translations can be
explained, to some extent, as the result of the Japanese listeners
accommodating the strings of English sounds to the Japanese phoneme
inventory and Japanese phonotactics
(\citeproc{ref-dupoux1999epenthetic}{Dupoux et al., 1999};
\citeproc{ref-peperkamp2008perceptual}{Peperkamp et al., 2008}). Native
prosody also constrains the occurence of homophonic translations. For
instance, Kentner (\citeproc{ref-kentner2015rhythmic}{2015}) presented
in the auditory modality French-native and German-native participants
with English songs with lyrics. Homophonic translations followed
language-specific segmentation strategies. Most homophonic translations
reported by German-native participants resulted from inserting phrase
and word boundaries \emph{before} stressed syllables (following the more
frequent trochaic patter in German). On the contrary, most homophonic
translations reported by French-native participants resulted from
inserting phrase and word boundaries \emph{after} stressed syllables,
following the more frequent iambic pattern in French.

While the previous studies have address the phonological aspects of
homophonic translation, little attention has been paid to the dynamics
of lexical activation and competition that underlay this phenomenon. In
the present work we investigated the interplay between phonological
similarity and phonological neighbourhood density to examine how
listeners of an unfamiliar language activate lexical representations in
their native language. In particular, we studied how the presence of
phonological neighbours in the native language lexicon affects the
extent to which listeners are able to activate the correct translation
of an auditorily presented word in an unfamiliar language. We used a
translation elicitation task in which participants listened to words
from an unfamiliar language, and then tried to guess their translation
in their native language. We will henceforth refer to the
auditory-presented words heard by participants on each trial as
presented words, and the correct translation for the presented words as
target translations. We explored listeners' reliance on phonological
similarity to succeed in the task by manipulating the amount of
phonological similarity between the presented words and their target
translations. Since participants were unfamiliar with the presented
language, they could only use phonological similarity between the
presented language and their native language to successfully translate
the words. We therefore predicted that participants' performance should
increase when the translation pairs are phonologically more similar. We
also predicted that there is a minimum threshold of phonological
similarity to be sufficient for translation.

We further investigated the effect of phonological competitors. Even if
the presented word and its target translation share high phonological
similarity, participants may still provide incorrect translations if the
presented word also shares high phonological similarity with other words
in the native language. Words with denser phonological neighbourhoods
are recognised more slowly and less accurately than words with sparser
phonological neighbourhoods (\citeproc{ref-dufour2003lexical}{Dufour \&
Peereman, 2003}; \citeproc{ref-goldinger1989priming}{Goldinger et al.,
1989}; \citeproc{ref-hamburger1996phonological}{Hamburger \& Slowiaczek,
1996}; \citeproc{ref-luce1990similarity}{Luce et al., 1990};
\citeproc{ref-luce1998recognizing}{Luce \& Pisoni, 1998}). This is
especially true if such neighbours share higher phonological similarity
with the presented word, or are lexically more frequent than the target
translation (\citeproc{ref-luce1998recognizing}{Luce \& Pisoni, 1998}).
Since the interference effects of phonological neighbourhood density has
also been reported across languages
(\citeproc{ref-weber2004lexical}{Weber \& Cutler, 2004}), we
hypothesised that the size of the facilitatory role of phonological
similarity would be inversely proportional to the amount of
higher-frequency phonological neighbours of the presented word in the
target language. In the case of non-native speech recognition, the
presence of cross-linguistic pairs which are phonologically similar but
differ in meaning (e.g., false friends) may act as distractors during
lexical access, obstructing the selection of the appropriate target
translation in the listener's lexicon. For instance, Otwinowska \&
Szewczyk (\citeproc{ref-otwinowska2019more}{2019}) reported that false
friends were disadvantaged relative to non-cognates by Polish second
language learners, in contrast to cognates which were known better. It
is therefore important to investigate the joint effect of both cognates
and false friends when investigating the effect cross-linguistic
phonological similarity has on word recognition in a foreign language.
This could shed light on available strategies and challenges associated
with the early stages of second language acquisition. For instance, one
could expect English participants to incorrectly translate the Spanish
word \emph{botón} as \emph{bottom} instead of as its correct translation
\emph{button}, due to combined effect of the close phonological
similarity between \emph{botón} and \emph{bottom} along with the high
lexical frequency of \emph{bottom} relative to \emph{button}. To test
this prediction, we developed a lexical frequency-dependent measure of
cross-language phonological neighbourhood density, in which a neighbour
is counted only if it is higher frequency and is one phoneme apart from
the presented word. If the phonological neighbourhood density of the
target translation affects participants' performance negatively, this
would suggest that competitors in the native language affect recognition
of non-native words during recognition of foreign speech. We conducted a
series of three experiments to test these predictions.

@@

This type of cross-linguistic similarity has also been extremely
informative for the study of bilingual lexical processing: words that
share high form-similarity with their translations are processed faster
and more accurately than words sharing lower phonological similarity.
This phenomenon has provided strong evidence in favour of the
language-non selective hypothesis of bilingual lexical access, which
states that bilinguals activate lexical representations in both
languages, even during monolingual situations. This facilitation effect
has been reported for comprehension
(\citeproc{ref-dijkstra2010cross}{Dijkstra et al., 2010};
\citeproc{ref-midgley2011effects}{Midgley et al., 2011};
\citeproc{ref-thierry2007brain}{Thierry \& Wu, 2007}), production
(\citeproc{ref-costa2000cognate}{Costa et al., 2000}), learning
(\citeproc{ref-de2000hard}{De Groot \& Keijzer, 2000};
\citeproc{ref-elias2022crosslanguage}{Elias \& Degani, 2022};
\citeproc{ref-lotto1998effects}{Lotto \& De Groot, 1998};
\citeproc{ref-valente2018does}{Valente et al., 2018}), and translation
(\citeproc{ref-christoffels2006memory}{Christoffels et al., 2006}).

@@

\bibitem[\citeproctext]{ref-dupoux1999epenthetic}
Dupoux, E., Kakehi, K., Hirose, Y., Pallier, C., \& Mehler, J. (1999).
Epenthetic vowels in japanese: A perceptual illusion? \emph{Journal of
Experimental Psychology: Human Perception and Performance},
\emph{25}(6), 1568.

@@

@@

\bibitem[\citeproctext]{ref-kentner2015rhythmic}
Kentner, G. (2015). Rhythmic segmentation in auditory illusions-evidence
from cross-linguistic mondegreens. \emph{ICPhS}.

@@

@@

\bibitem[\citeproctext]{ref-otake2007interlingual}
Otake, T. (2007). Interlingual near homophonic words and phrases in L2
listening: Evidence from misheard song lyrics. \emph{Proceedings of the
16th International Congress of Phonetic Sciences (ICPhS 2007)},
777--780.

@@

\bibitem[\citeproctext]{ref-peperkamp2008perceptual}
Peperkamp, S., Vendelin, I., \& Nakamura, K. (2008). On the perceptual
origin of loanword adaptations: Experimental evidence from japanese.
\emph{Phonology}, \emph{25}(1), 129--164.

@@
